---
title: "Lab 3: COVID-19"
subtitle: 'ESS 330 - Quantitative Reasoning'
author:
  - name: "Zachary Cramton"
    url: "https://zcramton.github.io"
    email: "ZCramton@colostate.edu"
format: html
toc: true
---

```{r}
# Load necessary libraries
library(tidyverse)
library(flextable)
library(zoo)
```

## Question 1: Public Data

```{r}
# Read in and store NY Times US county covid data
census_url <- "https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv"
us_covid_data <- read_csv(census_url)

# Read in and store US Census data
covid_url <- "https://www2.census.gov/programs-surveys/popest/datasets/2020-2023/counties/totals/co-est2023-alldata.csv"
us_census <- read_csv(covid_url)

# Check data structure to ensure it's loaded properly (remove the "#" on the line below to see output)
  #str(covid_data)
  #str(us_census)
```

DOTHIS: Text based reflection on the value of open source data.

## Question 2: Daily Summary
```{r}
# Define state and date for the analysis
state_name <- "Colorado"
target_date <- "2022-02-01"

# Convert date column to Date type
us_covid_data$date <- as.Date(us_covid_data$date)

# Create a subset limiting data to Colorado with new case and death information.
co_covid_data <- us_covid_data %>% 
  filter(state == state_name) %>%              # Filter Colorado data
  #summarize(county) %>%                       # Collate data for each county
  mutate(new_cases = cases - lag(cases),       # Create new column for new cases
         new_deaths = deaths - lag(deaths))    # Create new column for new deaths

# Create tables displaying the 5 counties with the highest cum/new cases.

# Counties with the worst cases
worst_cum_cases <- co_covid_data %>%
  filter(date == target_date) %>% 
  slice_max(cases, n = 5) %>% 
  select(-state, -fips)

#Counties with the worst deaths
worst_new_cases <- co_covid_data %>%
  filter(date == target_date) %>% 
  slice_max(new_cases, n = 5) %>% 
  select(-state, -fips)

# Print worst cumulative cases table
flextable(worst_cum_cases) %>%
  set_header_labels(
    date = "Date",
    county = "County",
    cases = "Cumulative Cases",
    deaths = "Cumulative Deaths",
    new_cases = "New Cases",
    new_deaths = "New Deaths"
  ) %>% 
  set_caption("Top 5 Colorado Counties for Cumulative Cases") %>% 
  align(part = "all", align = "center")

#Print worst new cases table
flextable(worst_new_cases) %>% 
    set_header_labels(
    date = "Date",
    county = "County",
    cases = "Cumulative Cases",
    deaths = "Cumulative Deaths",
    new_cases = "New Cases",
    new_deaths = "New Deaths"
  ) %>% 
  set_caption("Top 5 Colorado Counties for New Cases") %>% 
  align(part = "all", align = "center")
```
## Question 3: Normalizing Data
```{r}
# Reformat FIPS data from 2-3 to 5 digit combined strings.
us_census_formatted <- us_census %>% 
  mutate(
    STATE = sprintf("%02d", as.numeric(STATE)),     # Format state to 2 digits
    COUNTY = sprintf("%03d", as.numeric(COUNTY)),   # Format county to 3 digits
    fips = paste0(STATE, COUNTY)   # Combine state and county codes to get full 5 digit FIPS code.
  )

#Refine US Census Data
us_census_2021 <- us_census_formatted %>% 
  # Remove state level data DOTHIS: make   sure "000" for CO is being removed
  filter(COUNTY != "000") %>%    
  
  # Keep only columns with "NAME" or       "2021" (or the FIP Column)
  select(
         contains("NAME"),
         contains("2021"),
         fips)   
```

```{r}
# Explore Census and CO COVID Data
str(us_census_2021)
str(co_covid_data)
```
DOTHIS: In a few sentences describe the data & any similarities.

```{r}
# Find range of CO pops in 2021
# Filter for CO FIPS code (08)
co_pop <- us_census %>% 
  filter(STATE == "08") %>% 
  group_by(COUNTY)

# Calculate population range in CO in 2021
co_pop_range <- range(co_pop$`POPESTIMATE2021`, na.rm = TRUE)

#Print colorado pop range
co_pop_range
```
In 2021 Colorado's least populous county had 741 permanent residents and the most populous county had 5,811,596 permanent residents.
```{r}
# Join US Census data with CO covid data for 2021
co_combined <- co_covid_data %>% 
  left_join(us_census_2021, by = "fips") 

# Calculate per capita (pc) statistics
co_combined <- co_combined %>%
  rename(pop_2021 = POPESTIMATE2021) %>% 
  mutate(
    pc_cum_cases = cases / pop_2021,   # cumulative cases per capita
    pc_cum_deaths = deaths / pop_2021,  # cumulative deaths per capita
    pc_new_cases = new_cases / pop_2021,   # new cases per capita
    pc_new_deaths = new_deaths / pop_2021,  # new deaths per capita
  ) %>% 
  # Narrow df to remove superfluous columns from census      data
  select(1:8, 11, 27:30) #DOTHIS: Ask about more efficient way to execut; remove all with cap letters in header?

# Generate tables (2) for the 5 counties with highest cumulative and new cases par capita for a target date.

# Set new target date (if desired)
target_date <- target_date

# Counties with the worst cases
worst_pc_cum_cases <- co_combined %>%
  filter(date == target_date) %>% 
  slice_max(pc_cum_cases, n = 5) %>% 
  select(-state, -fips)

#Counties with the worst deaths
worst_pc_new_cases <- co_combined %>%
  filter(date == target_date) %>% 
  slice_max(pc_new_cases, n = 5) %>% 
  select(-state, -fips)

# Print worst cumulative cases table
flextable(worst_pc_cum_cases) %>%
  set_header_labels(
    date = "Date",
    county = "County",
    cases = "Cumulative Cases",
    deaths = "Cumulative Deaths",
    new_cases = "New Cases",
    new_deaths = "New Deaths",
    pc_cum_cases = "Cumulative Cases Per Capita",
    pc_cum_deaths = "Cumulative Deaths Per Capita",
    pc_new_cases = "New Cases Per Capita",
    pc_new_deaths = "New Deaths Per Capita"
  ) %>% 
  set_caption("Top 5 Colorado Counties for Cumulative Cases Per Capita") %>% 
  align(part = "all", align = "center")

#Print worst new cases table
flextable(worst_pc_new_cases) %>% 
    set_header_labels(
    date = "Date",
    county = "County",
    cases = "Cumulative Cases",
    deaths = "Cumulative Deaths",
    new_cases = "New Cases",
    new_deaths = "New Deaths",
    pc_cum_cases = "Cumulative Cases Per Capita",
    pc_cum_deaths = "Cumulative Deaths Per Capita",
    pc_new_cases = "New Cases Per Capita",
    pc_new_deaths = "New Deaths Per Capita"
  ) %>% 
  set_caption("Top 5 Colorado Counties for New Cases Per Capita") %>% 
  align(part = "all", align = "center")

```

## Quesiton 4: Rolling Thresholds

```{r}
# Get the latest date in the data set
latest_date <- max(co_combined$date, na.rm = TRUE)

# Filter for the last 14 days
two_week_data <- co_combined %>% 
  filter(date >= (latest_date - 14))

# Summarize new cases per 100,000 residents
summarized_per_100k <- two_week_data %>% 
  group_by(county) %>% 
  summarize(total_new_cases = sum(new_cases, TRUE),
            population = first(population)) %>% 
  mutate(cases_per_100k = (total_new_cases / population) * 100000) %>% 
  arrange(desc(cases_per_100k))

# Print the 5 worst colorado counties for 14-day new case numbers
worst_14_day_counties <- head(summarized_per_100k, 5)
print(worst_14_day_counties)

# Count counties meeting the watch list condition (>100 cases per 100,000)
watchlist_count <- worst_14_day_counties %>% 
  filter(cases_per_100k > 100) %>%
  nrow()

print("Number of Colorado counties meeting the watch list condition:", watchlist_count, "\n")

```


## Question 5: Death toll


## Question 6: Multi-State


## Quesiton 7: Space & Time


## Question 8: Cases vs. Deaths